One exact and one random sampling method to assess the reliability of inferences from network enrichment analysis

### 2018-08-01 doc/genesetdp.tex
GEA -> ORA after talking to Andrey

% Use the asterisk to denote corresponding authorship and provide email address in note below.
*lukas.kall@scilifelab.se
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
A prevailing technique to infer function from identifications, from molecular biological high-throughput experiments, is gene set enrichment analysis, where the identifications are compared to predefined sets of related genes often refered to as pathways. As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks. While the termonolgy vary in the litterature, we will here refer to such methods as Network Enrichment Analysis (NEA). Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations. Here we instead have designed one dynamic programming  algorithm that calculates the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences. We also implemented a random sampling method, carrying out the same task. We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engine, BinoX.

@@

Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts, proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten,goeman2007analyzing}. The technique estimates which functional modules, such as complexes or pathways, are overrepresented among a set of identified analytes. One prominent application of the technique is expression analysis, where GEA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.

Most GEA methods are assessing the overlap between the investigated set of analytes, the {\em query set}, and a functional module, the {\em pathway set}, using hypergeometric test or a Fisher's exact test. However, variants such as Gene Set Enrichment Analysis (GSEA)\cite{subramanian2005gene} also includes information on expression levels of the analytes of the query set.

A limiting factor of GEA is the pathway definition databases. While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases. Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string} and FunCoup\cite{ogris2017funcoup}. Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}. While the termonolgy vary in the litterature, we will here refer to such methods as {\em Network Enrichment Analysis} (NEA).

>>>>

% Use the asterisk to denote corresponding authorship and provide email address in note below.
*lukas.kall@scilifelab.se
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
A prevailing technique to infer function from lists of identifications, from molecular biological high-throughput experiments, is over-representation analysis, where the identifications are compared to predefined sets of related genes often refered to as pathways. As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks. While the termonolgy vary in the litterature, we will here refer to such methods as Network Enrichment Analysis (NEA). Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations. Here we instead have designed one dynamic programming  algorithm that calculates the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences. We also implemented a random sampling method, carrying out the same task. We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engine, BinoX.

@@

Over-Representation Analysis (ORA) is commonly used to infer function from sets of analytes such as genes, transcripts, proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten,goeman2007analyzing}. The technique estimates which functional modules, such as complexes or pathways, are overrepresented among a set of identified analytes. One prominent application of the technique is expression analysis, where ORA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.

Most ORA methods are assessing the overlap between the investigated set of analytes, the {\em query set}, and a functional module, the {\em pathway set}, using hypergeometric test or a Fisher's exact test. However, variants such as Gene Set Enrichment Analysis (GSEA)\cite{subramanian2005gene} also includes information on expression levels of the analytes of the query set.

A limiting factor of ORA is the pathway definition databases. While \url{http://pathguide.org} list 708 databases of pathway definitions\cite{bader2006pathguide},  some critique has been voiced concerning the completeness and rigor of current pathway databases. Hence efforts have been directed to designing methods that extend the pathway definitions, by functional association networks, like STRING\cite{szklarczyk2014string} and FunCoup\cite{ogris2017funcoup}. Instead of directly examining the overlap between the query and pathway sets, one can evaluate the number of links in the functional association network that connect the query and pathway set\cite{alexeyenko2012network, glaab2012enrichnet, mccormack2013statistical, ogris2016novel, signorelli2016neat}. While the termonolgy vary in the litterature, we will here refer to such methods as {\em Network Enrichment Analysis} (NEA).
